---
title: "Pharmacometric simulation with rxode2"
description: "Basically the same post as the last one, but this time using rxode2 instead of mrgsolve"
date: "2023-08-20"
--- 

<!--------------- my typical setup ----------------->

```{r}
#| label: setup
#| include: false
very_wide <- 500
wide <- 136
narrow <- 76
options(width = narrow)
cache_images <- TRUE
set.seed(1)

# enable ansi-to-html colouring
options(crayon.enabled = TRUE)
old_hook <- fansi::set_knit_hooks(
  knitr::knit_hooks, 
  which = c("output", "warning", "error", "message")
)
```

<!--------------- post begins here ----------------->

This post is about the [rxode2](https://nlmixr2.github.io/rxode2/) package, the successor to [RxODE](https://nlmixrdevelopment.github.io/RxODE/). Although the original RxODE package is now archived on CRAN, the syntax for rxode2 is very similar, and as far as I can tell it's fully backward-comparible with the older package. When getting started, I found it a little easier to look at the [rxode2 user manual](https://nlmixr2.github.io/rxode2-manual/) than to work from the pkgdown site. 


## Installation

As with other packages for pharmacometric simulation such as [mrgsolve](https://mrgsolve.org/) -- which I wrote about in my [last blog post](/posts/2023-08-14_mrgsolve/) -- models defined with rxode2 need to be compiled before they are run, and so when you install the package you need the appropriate build tools. There are some implications to this. The package is on CRAN, so you can install it with:

``` r
install.packages("rxode2")
```

However, like most R packages that allow you to compile C/C++/Fortran/Rust/Your-Favourite-Language-Here code, it relies heavily on system dependencies that you may or may not have, and managing the build tools is an OS-specific thing. I'm running Ubuntu 22.04, and (for reasons that don't bear mentioning) I recently did a "factory reset"^[Does that term even make sense for a linux machine? It's not like the thing shipped with linux in the first place. Whatever.] and did a fresh install of Ubuntu. So, yeah, I didn't have everything I needed. Yes, I did have the `gcc` compiler installed, but that's not the only system dependency you have to care about. In my case, I was missing `gfortran`, `libblas`, and `liblapack`. As a consequence, when I tried to run the example code on the package website, all I got was a long stream of error messages. In order to get started, I had to do this: 

``` bash
sudo apt install gfortran libblas-dev liblapack-dev liblapack-doc
```

That worked for me.^[I haven't extensively checked the dependencies on other operating systems, but from what I can tell a Windows install requires [RTools](https://cran.r-project.org/bin/windows/Rtools/).]

```{r}
library(rxode2)
```

## The rxode2 mini-language

The rxode2 package uses a [domain-specific language](https://nlmixr2.github.io/rxode2/articles/rxode2-syntax.html) for specifying pharmacometric models. You can define an rxode2 model in a separate file or you can do it inline by passing a string, or enclosing the model specification within an R code block (i.e., enclosed in curly braces `{}`). I have to admit that most of the time when an R package does this, I have a strong preference to write a separate model specification file. That's what I do when building [Stan](https://mc-stan.org/) models, and it's what I did with mrgsolve in the previous post. But this time around I'll use the curly braces method:

```{r}
mod <- rxode2({
  # initial values for all four "compartments"
  depot(0) = 0;
  central(0) = 0;
  peripheral(0) = 0;
  auc(0) = 0;
  
  # drug concentrations
  CP = central / VC;    # central compartment concentration
  PP = peripheral / VP; # peripheral compartment concentration
  
  # differential equations
  d/dt(depot)       = -(KA * depot);
  d/dt(central)     =  (KA * depot) - (Q * CP) + (Q * PP) - (CL * CP);
  d/dt(peripheral)  =  (Q * CP) - (Q * PP);
  d/dt(auc)         =  CP;
})
```

As usual, there's a few things that we need to unpack here. 

Firstly, note the message printed to the console. When an rxode2 model is built, this model specification is translated to C code and compiled: this message lets you know what C compiler rxode2 is using for that purpose.

Secondly, notice that although this is a two-compartment model in pharmacometric terms, from the perspective of rxode2 there are four "compartments". In addition to the usual two compartments (`central` and `peripheral`), there is an extravascular `depot` compartment used to model drug intake. For instance, for an orally-administered drug the `depot` compartment would be the gut.^[In this post I'm assuming the drug has bioavailability of $F = 1$, but that's not true generally, so you'd have to model this explicitly by scaling the drug amount that passes from the gut to the central compartment in the ODE equations.] The `depot` compartment is "real" in the sense that it is loosely intended to correspond to something in the physical system that we're modelling. By convention we don't consider it to be one of the pharmacokinetic compartments, but it's still a real thing. In contrast, the `auc` "compartment" has no physical analog at all. It's included so that the model keeps track of the accumulated drug exposure.^[In essence, the value of `auc` that accrues is a numerical estimate of the time-integral of drug concentration. This "area under the curve" measure is one of several different measures used to assess drug exposure. I talked a lot about the AUC measure in my post on [non-compartmental analysis](/posts/2023-04-26_non-compartmental-analysis/).] As I'm quickly coming to learn, this is a very handy trick when running pharmacometric simulations.

Finally, notice that the syntax within the rxode2 model specification is "R-like" rather than actual R syntax. Thanks to the magic of non-standard evaluation in R, the `rxode2()` function can capture the unevaluated expression I passed within the curly braces, and then parse the statements according to the rxode2 mini-language. The mini-language is very similar to R in some ways:

- Assignment statements can use `=` or `<-` as the assignment operator.^[The rxode2 mini-language also allows you to use `~` for this purpose, but I'm not going to do that here. For this post, I've chose to use `=` as a way of reminding myself that my model specification isn't "normal" R code.]
- Comments are specified using the hash (`#`) character
- Semi-colon characters (`;`) are optional, and specify the end of a line

There are some other specialised statements that don't exist in regular R code. There are two kinds of special statements I've used in this code:

- Time-derivative statements (i.e., the ones that have something like `d/dt(central)` on the left hand side) are used to specify the differential equations in the ODE system.
- Initial-condition statements (i.e., the ones where I set something like `central(0)` on the left hand side) are used to specify the initial state of the ODE system. 

You can check the [rxode2 syntax](https://nlmixr2.github.io/rxode2/articles/rxode2-syntax.html) page to see what other kinds of special statements exist.

## The rxode2 model object

In the previous section I used the `rxode2()` function to specify a pretty standard two-compartment pharmacokinetic model, and assigned the resulting model object to a boringly-named variable called `mod`.^[Model objects in rxode2 have S3 class "rxode2".] Let's take a look at it:

```{r}
mod
```

Again, a few things to unpack: 

- The first line of the output has some technical information about the model. It tells us what version of rxode2 was used to build the model, gives us the name of the built model (see below), and tells us that it's ready to use.^[You can customise this name if you care deeply about such things. As noted in the [`roxde2()` documentation](https://nlmixr2.github.io/rxode2/reference/rxode2.html), there is a `modName` argument that you can use for this purpose. Because this name is used throughout the C compilation process, it must start with a letter and contain only alphanumeric ASCII characters.]
- The second line tells us about `mod$state`, which in this case are the four "compartment" variables that comprise the state vector for the underlying ODE system.
- The third line tells us about `mod$params`, the list of parameters that need to be passed to the model as input to the simulation
- The fourth line tells us about `mod$lhs`, the list of additional defined variables that are created by the model and whose value will be recorded in the output.

Like many R packages that generate compiled code, rxode2 manages the compiled object for you. The long unintelligible "name" assigned to our model gives us the hint we need to find the compiled objects. Within the R session temp directory, the rxode2 package has created an "rxode2" subfolder.^[Yes, you can customise this too, by specifying the `wd` argument to `rxode2()`.] And indeed, if I take a peek at the contents of this folder, I find something with an identical name:

```{r}
fs::dir_ls(fs::path(tempdir(), "rxode2"))
```

Okay, makes sense. 

## Event tables

Event tables (also called event schedules) in the rxode2 package are specified with the `et()` function, and you can use the pipe operator to build up complex event schedules. I'll take my example from the rxode2 documentation, and walk through it slowly. One nice thing about the event schedules in rxode2 is that you can specify units, so we'll start with an event table that doesn't contain any actual *events*, but specifies the units in which those events will be expressed: 

```{r}
events <- et(amountUnits = "mg", timeUnits = "hours")
events
```

The output here isn't super exciting, since there are no actual events encoded here. But it does let me mention one nice little feature of rxode2: the print methods are generally quite informative, and have nice little "nudges" like the ones you can see above that can help new (or even experienced) users work out what they might need to do next. 

Anyway, let's add some dosing events, shall we? Let's assume an initial dose of `amt = 10000` (in milligrams) is administered at `time = 0`, and repeated for an additional 9 times at 12 hour intervals (i.e., `addl = 9`, `ii = 12`). In the interests of being explicit, I'll set `cmt = "depot"` to be clear about which compartment the dose is administered to. 

```{r}
events <- events |> 
  et(time = 0, amt = 10000, addl = 9, ii = 12, cmt = "depot")

events
```

This format for an event table -- where `time`, `amt`, `addl`, and `ii` are used to specify a sequence of regularly spaced dosing events in a single row -- will seem quite familiar to anyone in the field, and since I've talked about this notation in previous posts, I'll not bore people by explaining it yet again. 

Moving along, let's also assume that after 120 hours has passed (`time = 120`) the dosing schedule changes: the dose drops to `amt = 2000` milligrams, the interdose interval is increased slightly to `ii = 14` hours, and this dosing regime is maintained for `addl = 4` additional doses (i.e., 5 in total). So now we have this:

```{r}
events <- events |> 
  et(time = 120, amt = 2000, addl = 4, ii = 14, cmt = "depot")

events
```

Now that we have specified all the dosing events, we need to add the "observation" events. In a real study, observation times would be the times at which we take a real-world measurement of some kind, but in the context of the simulation it's just a set of times at which the state of the system is computed. Let's compute the state of the system for the first 300 hours:

```{r}
events <- events |> et(time = 0:300)
events 
```

And now we're done. We have a complete events table that can be used in our simulation. Admittedly, I went through that awfully slowly. The whole thing could have been bundled into a single pipeline like this:

```{r}
events <- et(amountUnits = "mg", timeUnits = "hours") |>
  et(time = 0, amt = 10000, addl = 9, ii = 12, cmt = "depot") |>
  et(time = 120, amt = 2000, addl = 4, ii = 14, cmt = "depot") |>
  et(time = 0:300)
```

## Simulating one subject

```{r}
library(tibble)
library(ggplot2)
```

We're now almost at a point where we can run a simple simulation using the model specified via the `mod` object, and the events table in `events`. The only thing we haven't done yet is specify input parameters. The model specification I wrote at the start of the post relies on pharmacokinetic parameters. If we look at the model spec we can see that requires all of the following to be given as input:

- elimination clearance (`CL`)
- absorption rate constant (`KA`)
- intercompartmental clearance (`Q`)
- volume of distribution for the central compartment (`VC`)
- volume of distribution for the peripheral compartment (`VP`)

Indeed, if we take a look at `mod$params` we see the same listing:

```{r}
mod$params
```

Okay, so let's put together a one-row data frame `params` containing all these parameters for a single simulated subject:

```{r}
params <- tibble(
  KA = 0.294,
  CL = 18.6,
  VC = 40.2,
  VP = 297,
  Q = 10.5
)
params
```

Now that we have our parameters, we're ready to go. There are several ways you can call the solver and run the simulation ([documentation here](https://nlmixr2.github.io/rxode2/reference/rxSolve.html)), but I'm currently quite partial to calling `solve()`,^[Experienced R users would not be surprised to discover that `solve()` is an S3 generic defined in the base package, and equally unsurprised to note that rxode2 defines a method for "rxode2" objects such as `mod`. It somehow makes me happy to see `solve()` used this way.] like so:

```{r}
out <- solve(mod, params, events)
```

When we print `out`, we get a fairly detailed description of the simulation that includes information about the parameters and the initial state: 

```{r}
out
```

Extremely pretty print method notwithstanding, under the hood it's nothing fancy. It's a regular data frame with a few extra classes and some metadata, which means we can pass it straight to ggplot without any coercion, and draw a pretty picture:

```{r}
ggplot(out, aes(time, CP)) + 
  geom_area(linewidth = 1) + 
  ylab("concentration") +
  theme_bw()
```


## Simulating multiple subjects

```{r}
params <- tibble(
  KA = rnorm(20, mean = 0.294, sd = 0.03),
  CL = rnorm(20, mean = 18.6, sd = 2),
  VC = rnorm(20, mean = 40.2, sd = 2),
  VP = rnorm(20, mean = 297, sd = 10),
  Q = rnorm(20, mean = 10.5, sd = 1)
)

out <- solve(mod, params, events)
```

I'll show you the `out` object in a moment, but it's probably easier to understand it if we start with a plot:

```{r}
#| fig-height: 10
out |>
  dplyr::mutate(sim.id = paste("Subject", sim.id)) |>
  ggplot(aes(time, CP, fill = sim.id)) + 
  geom_area(linewidth = 1, show.legend = FALSE) + 
  facet_wrap(~ sim.id, nrow = 5) + 
  labs(x = "Time", y = "Concentration") +
  theme_bw()
```

As you can see, all 20 subjects have qualitatively similar profiles, but there are noticeable differences in the details. Not surprisingly really. I didn't build in very much variability into the simulation, and I didn't even *try* to incorporate an appropriate covariance structure among the parameters (that's a topic for another post). The main thing that matters here is that we can see that the variation *exists*. 

Anyway, let's have a look at the table of results `out` produced by our simulation. As you probably guessed from the ggplot2 code, there's a column called `sim.id` that stores the subject identifier, and there are 20 times as many rows as last time, but it's essentially the same:

```{r}
out
```

As we've seen throughout the post, the print method has lots of nice touches. It shows the simulation parameters as well as the simulation results, and has a very gentle message reminding me I haven't incorporated measurement error, random effects, or parameter uncertainty. Which... I mean, I intentionally left those things out, but actually I do appreciate the clear statement of what *wasn't* done here. 

